A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-10 paper/paper.tex
Minor

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[nonacm]{acmart}
\usepackage{ottalt}
\usepackage{xspace}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

The consistency proofs of various dependently typed systems, including Martin-Lof's
type theory and the Calculus of Constructions, have long been
available in the literature. In particular, recent works such as \citet{nbeincoq},
\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
results related to the decidability of type conversion or type
checking of dependently typed systems, from which consistency can be
derived as a corollary. The underlying technique of the forementioned
works is proof by logical relation,
which involves interpreting types as reducibility predicates,
representing sets of terms satisfying certain properties with respect
to the reduction relation. While the proof technique and the
consistency result for dependent types are both well-established,
there is a severe lack of rigorous and accessible material that shows
how proof by logical relation can be applied to dependently typed
systems.
Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages, and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity. \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. In short, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms. The glaring gap here is the
lack of fully dependently typed systems where computations may appear
at the type-level. In fact, it is far from obvious why proof by
logical relation is even applicable to dependent types, since the type
may very well-be a computation that is yet to be evaluated.
While it is assuring that proof by logical relations for dependent
types is available in mechanized forms, % the key to address the
% complexities of dependent types is obscured in
\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
Kripke-style models that obscure the underlying technique of applying
proof by logical relation to dependent types. In fact, the
complexivity is self-evident from the size of their code base. For
example, \citet{nbeincoq} involves 22 thousand lines of Coq code.

@@

Our consistency proof is short and fully mechanized. The proof scripts
involve less than 1000 lines of Coq code. Furthermore, the mechanized
proof closely corresponds to the informal

@@

This paper is specifically
about establishing logical consistency for a fully dependently typed
system with an infinite universe hierarchy and support for large
elimination . % The type system, presented in Section~\ref{sec:spec}, is
% most similar to Martin-Lof's predicate type theory with the minor
% difference that type conversion is based on untyped equality.

Rather,
our goal is to present the proof in a form that is digestable by a
working type theorist and can be more readily mechanized in a proof
assistant. Compared to existing efforts at mechanizing logical
consistency or stronger properties such as existence of normal
form~\citep{nbeincoq},
decidable type checking~\citep{decagda}, our work is minimal since it requires very
little scaffolding and therefore results in an extremely succinct
proof of under 1000 lines of manually written Coq code for a dependent
type theory that is reasonably complete in its features.

The key technique that underlies our consistency proof is proof by
logical relation. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
which states that syntactic typing implies semantic typing. Once the
fundamental theorem is established, logical consistency follows as a
trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
specifics related to the Coq mechanization of the proof described in earlier
section.
Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
existing literature related to logical consistency about dependent
type theory.

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
the unscoped DeBruijn representation to leave less ambiguity for the
formulation of the system and simplify the reasoning about binders.
Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
syntax, including renaming, substitution, and operations over
the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{ottalt}
\usepackage{xspace}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

The consistency proofs of various dependently typed systems, including Martin-Lof's
type theory and the Calculus of Constructions, have long been
available in the literature. In particular, recent works such as \citet{nbeincoq},
\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
results related to the decidability of type conversion or type
checking of dependently typed systems, from which consistency can be
derived as a corollary.

The underlying technique of the forementioned
works is proof by logical relation,
which involves interpreting types as reducibility predicates,
representing sets of terms satisfying certain properties with respect
to the reduction relation. While the proof technique and the
consistency result for dependent types are both well-established,
there is a severe lack of rigorous and accessible material that shows
how proof by logical relation can be applied to dependently typed
systems.

Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages, and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity. \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. In short, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms.

The glaring gap here is the
lack of fully dependently typed systems where computations may appear
at the type-level. In fact, it is far from obvious why proof by
logical relation is even applicable to dependent types, since the type
may very well-be a computation that is yet to be evaluated.
While it is assuring that proof by logical relations for dependent
types is available in mechanized forms, % the key to address the
% complexities of dependent types is obscured in
\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
Kripke-style models that obscure the technique for addressing
type-level computation. In fact, the added complexivity is evident
from the size of their code base. For example, \citet{nbeincoq}
involves 22 thousand lines of Coq code.

@@

Our consistency proof is short and fully mechanized. The proof scripts
involve less than 1000 lines of Coq code. In fact, what is even more
encouraging is that among the 1000
lines of Coq code, 400 lines are related to the specification of the
type system, semantics, and properties related to untyped lambda
terms. The semantic type soundness proof through logical relation
takes almost the same amount of code as our syntactic type soundness
proof!
Thanks to the conciseness of
the proof, we are able to present most of its details in
Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
structure of our mechanization closely corresponds to the proof we
present in the text, enabling us to label at the footnote each lemma
directly to their counterpart in the proof script for the readers to
reference and validate.

The technique we use is most similar to the one from \citet{nbeincoq},
which leverages impredicativity to define the logical
relation. % In our presented proof, we use the informal
% language of sets, where the use of impredicativity feels more natural
% and less noticeable.
Rather than framing impredicativity as a
mechanism for encoding induction
recursion~\citep{induction-recursion}, a scheme in which logical
relation for dependent types is typically defined~\citep{todo}, we opt
for a direct explanation in informal language of sets, where
impredicativity manifests in the form of second-order logical
formulas and thus more intuitive to understand for readers with some
general mathematics background.

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation. In Section~\ref{sec:discuss}, we show that extensions such
as $\eta$ rules for functions and normalization for open terms are
more or less orthogonal to the fact that the language is dependently
typed. In our mechanized proof of said extensions, the overall
structure of the logical relation is preserved, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

We hope our success at creating a short and mechanized proof for a
relatively feature-complete language will encourage future researchers
to leverage the tool of logical relation more often in mechanized
proofs for dependent types.

@@

% This paper is specifically
% about establishing logical consistency for a fully dependently typed
% system with an infinite universe hierarchy and support for large
% elimination .
% The type system, presented in Section~\ref{sec:spec}, is
% most similar to Martin-Lof's predicate type theory with the minor
% difference that type conversion is based on untyped equality.

% Rather,
% our goal is to present the proof in a form that is digestable by a
% working type theorist and can be more readily mechanized in a proof
% assistant. Compared to existing efforts at mechanizing logical
% consistency or stronger properties such as existence of normal
% form~\citep{nbeincoq},
% decidable type checking~\citep{decagda}, our work is minimal since it requires very
% little scaffolding and therefore results in an extremely succinct
% proof of under 1000 lines of manually written Coq code for a dependent
% type theory that is reasonably complete in its features.

% The key technique that underlies our consistency proof is proof by
% logical relation. In
% Section~\ref{sec:spec}, we present the dependent type theory of
% interest. In Section~\ref{sec:logreldep}, we give the definition
% of the logical relation for the dependent type theory. Rather than
% presenting the logical relation as an inductive-recursive definition,
% we use the more elementary concept of a partial function to capture
% the interpretation of types. The alternative representation requires us
% to show that the set of equations indeed defines a partial function;
% that is, for each input, there should always be a unique
% output.
% From the interpretation function, we can define the semantic
% typing judgment for the set of lambda terms.
% In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
% which states that syntactic typing implies semantic typing. Once the
% fundamental theorem is established, logical consistency follows as a
% trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
% specifics related to the Coq mechanization of the proof described in earlier
% section.
% Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
% existing literature related to logical consistency about dependent
% type theory.

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. To help
the reader understand the binding structure of the terms, we annotate
the terms with their DeBruijn depth. For example, ....
 We use
the unscoped DeBruijn representation to leave less ambiguity for the
formulation of the system and simplify the reasoning about binders.
Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
syntax, including renaming, substitution, and operations over
the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.
